Local Spectral Clustering of Density Upper Level Sets

### 2019-01-23 icml/icml_local_density_clustering.tex
Edits through Sect 2

\begin{abstract}
\vskip 0.1 in % RJT: there was no space here and it looked very squished 
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several local methods, such as the personalized PageRank (PPR)
algorithm, have been shown to have worst-case guarantees for certain graph-based
measures of cluster quality.  In contrast to existing works, we analyze PPR in a
traditional statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.   
\end{abstract}

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be been computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,anderson2006,mahoney2012,zhu2013}, among 
others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,mahoney2012,REFS}, which
has led many authors to develop theory that explains their success 
\citep{zhu2013,REFS},
from the perspective of worst-case guarantees on traditional graph-theoretic
notions of cluster quality (like conductance).  In this paper, we adopt a more
traditional statistical viewpoint, and examine what the output of a local
clustering algorithm on $\Xbf$ reveals about the unknown density $f$.  
In particular, we examine the ability of the PPR algorithm to recover
\emph{density clusters} of $f$, which are defined as the connected components of
the upper level set $\{x : f(x) \geq \lambda\}$ for some threshold $\lambda >
0$ (a central object of central interest in the classical statistical literature
on clustering, dating back to \citealt{hartigan1981}).

@@

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no
less than $\Psi$, we call it as a \emph{$(\Phi,\Psi)$-cluster}. Both
local \cite{zhu2013} and global \cite{kannan04} spectral algorithms have been
shown to output clusters (or partitions) which approximate the optimal $(\Phi,
\Psi)$-cluster (or partition) for a given graph $G$.\footnote{In the case of
  \cite{kannan04}, the internal connectivity parameter $\phi$ is actually the
  conductance, i.e., the minimum normalized cut within the subgraph $G[S]$. See
  Theorem 3.1 in their paper for details; however, note that $\phi^2 /
  \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a
  lower bound on $\Psi$.}

@@

For a level $\beta > 0$ and a target volume $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of \pprspace
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target volume $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return $\widehat{C} = S_{\beta^*}$, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

@@

%%% RJT: This was already commented out before

Let $\Cbb_f(\lambda)$ denote the connected components of the density upper level
set $\{x : f(x) > \lambda\}$.  For a given density cluster $\Cset \in
\Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical
  density cluster}. Below we define a notion of consistency in density cluster
estimation.

@@

\begin{enumerate}
\item In Section \ref{sec: PhiPsi}, we derive in Theorem \ref{thm:
    conductance_upper_bound} an upper bound on the normalized cut of an
  empirical density cluster $\Cset[\Xbf]$, under a natural set of    
  geometric conditions (precluding clusters which are too thin and long).

\item Under largely the same set of geometric conditions, we derive in Theorem
  \ref{thm: inverse_mixing_time_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Cset[\Xbf]$.  We also provide in
  Theorem \ref{thm: inverse_mixing_time_lower_bound} a tighter lower bound, but
  under more restrictive assumptions (convexity of $\Cset[\Xbf]$).

@@

\section{Cluster Quality Criteria on Empirical Density Clusters} 
\label{sec: PhiPsi}

In order to provide meaningful bounds on the normalized cut and inverse mixing time of an empirical density cluster $\Cset[\Xbf]$, we must introduce some assumptions on the density $f$.

Let $B(x,r) = \set{y \in \Rd: \norm{y - x} \leq r}$ be a closed ball of radius $r$ around the point $x$.  Given a set $\Aset \subseteq \Rd$, and a number $\sigma > 0$, define the $\sigma$-expansion of $\Aset$ to be $\Asig = \Aset + B(0,\sigma) = \set{y \in \Rd: \inf_{x \in \Aset} \norm{y - x} \leq \sigma}$. We are now ready to give the assumptions, which we state with respect to a density cluster $\Cset \in \Cbb_f(\lambda)$ for some $\lambda > 0$, and expansion parameter $\sigma > 0$:
\begin{enumerate}[label=(A\arabic*)]
	\item
	\label{asmp: cluster_separation}
	\textit{Cluster separation:}
	For all $\Cset' \in \Cbb_f(\lambda)$,
	\begin{equation*}
	\dist(\Csig,\Csig') > \sigma,
	\end{equation*}
	where $\dist(\Aset,\Aset') = \min_{x \in \Aset} \dist(x,\Aset')$ for $\Aset' \subseteq \Rd$.

	\item
	\label{asmp: cluster_diameter}
	\textit{Cluster diameter:}
	There exists $D < \infty$ such that for all $x, x' \in \Csig$:
	\begin{equation*}
	\norm{x - x'} \leq D.
	\end{equation*}

	\item
	\label{asmp: bounded_density}
	\textit{Bounded density within cluster:} There exist numbers $0 < \lambda_{\sigma} <  \Lambda_{\sigma} < \infty$ such that:
	\begin{equation}
	\label{eqn: bounded_density}
	\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq \Lambda_{\sigma} 
	\end{equation} 
	%and 
	%\begin{equation*}
	%\frac{\diam \Asig}{\sigma} \leq \mu
	%\end{equation*}
	%where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

	\item 
	\label{asmp: low_noise_density}
	\textit{Low noise density:} For some $\gamma > 0 $, there exists a constant $c_1 > 0$ such that for all $x \in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,
	\begin{equation*}
	\inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \dist(x, \Csig)^{\gamma},
	\end{equation*}
	where $\dist(x,\Aset) = \min_{x_0 \in \Aset} \norm{x - x_0}$ for $\Aset \subseteq \Rd$.
\end{enumerate}

We note that $\sigma$ plays several roles here, precluding arbitrarily narrow clusters and long clusters in \ref{asmp: cluster_diameter} and \ref{asmp: bounded_density}, flat densities around the level set in \ref{asmp: low_noise_density}, and poorly separated clusters in \ref{asmp: cluster_separation}.

Assumptions \ref{asmp: cluster_separation}, \ref{asmp: bounded_density} and \ref{asmp: low_noise_density} are used to upper bound $\Phi(\Cset[\Xbf]; G_{n,r})$, whereas \ref{asmp: cluster_diameter} and \ref{asmp: bounded_density} are necessary to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the lower bound on minimum density in (\ref{eqn: bounded_density}) and \ref{asmp: cluster_separation} combined are similar to the $(\sigma,\epsilon)$-saliency of \cite{chaudhuri2010}, a standard density clustering assumption, while \ref{asmp: low_noise_density} is seen in, for instance, \cite{singh2009}, (as well as many other works on density clustering and level set estimation.) It is worth highlighting that these assumptions are all local in nature, a benefit of studying a local algorithm such as PPR.

We are ready to provide bounds on the graph quality bicriteria. For notational simplicity, hereafter for $S \subseteq \Xbf$ we will refer to $\Phi(S; G_{n,r})$ as $\Phi_{n,r}(S)$, and likewise with $\Psi(S; G_{n,r})$ and $\Psi_{n,r}(S)$. We will also use $\nu(\cdot)$ to denote the uniform measure over $\Rd$, and $\nu_d = \nu\bigl(B(0,d)\bigr)$ as the measure of the unit ball.

We begin with an upper bound on the normalized cut in Theorem \ref{thm: conductance_upper_bound}. We will require Assumptions \ref{asmp: cluster_separation}, \ref{asmp: bounded_density} and \ref{asmp: low_noise_density} to hold; however, the upper bound on density in (\ref{eqn: bounded_density}) will not be needed and so we omit the parameter $\Lambda_{\sigma}$ from the statement of the theorem.
\begin{theorem}
	\label{thm: conductance_upper_bound}
	For some $\lambda > 0$, let $\Cset \in \mathbb{C}_f(\lambda)$ satisfy Assumptions \ref{asmp: cluster_separation}, \ref{asmp: bounded_density} and \ref{asmp: low_noise_density} for some $\sigma, \lambda_{\sigma}, c_1, \gamma > 0$ . Then, for any $r < \sigma/4d$ and $\delta \in (0,1]$, the following statements hold with probability at least $1 - \delta$:  Fix $\epsilon > 0$. Then, for
		\begin{equation}
		\label{eqn: conductance_sample_complexity}
		n \geq \frac{9\log(2/\delta)}{\epsilon^2}\left(\frac{1}{ \lambda_{\sigma}^2 \nu(\Csig) \nu_d r^d}\right)^2 
		\end{equation}
		we have
		\begin{equation}
		\label{eqn: conductance_additive_error_bound}
		\frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq 4 c_{\sigma} d \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} - \frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon
		\end{equation}
	where $c_{\sigma} = 1 / \sigma$. 
\end{theorem}

\begin{remark}
	The proof of Theorem \ref{thm: conductance_upper_bound}, along with all other proofs, can be found in the supplementary document. The key point is to note that for any $x \in \Cset$, the simple, (possibly loose) $B(x,\sigma) \subseteq \Csig$ translates to the upper bound $\nu(\Cset_{\sigma + r}) \leq (1 + 2d r /\sigma)\nu(\Csig)$. We leverage \ref{asmp: low_noise_density} to find a corresponding bound on the weighted volume, before applying standard concentration inequalities to convert from population to sample based results.
\end{remark}
\begin{remark}
	(\ref{eqn: conductance_additive_error_bound}) is almost tight. Specifically, choosing
	\begin{align*}
	\Asig & = B(0,\sigma), \\
	 f(x) & = 
	 \begin{cases}
	 \lambda \text{ for $x \in \Asig$}, \\
	 \lambda - \dist(x,\Asig)^{\gamma} \text{ for $0 < \dist(x,\Asig) < r$}
	 \end{cases}
	\end{align*}
	we have that for $n$ within constant order of the lower bound in (\ref{eqn: conductance_sample_complexity}), with probability at least $1 - \delta$
	\begin{equation*}
	\frac{\Phi_{n,r}(\Asig[\mathbf{X}])}{r} \geq c \frac{(\lambda - \frac{r^{\epsilon+1}}{\epsilon+1})}{\lambda} - \epsilon
	\end{equation*}
	for some constant $c$. (Note that a factor of $1 / \sigma$ in $c_{\sigma}$ is not replicated in this lower bound.)
\end{remark}

We now provide a lower bound on $\Psi_{n,r}(\Cset[\Xbf])$.

\begin{theorem}
	\label{thm: inverse_mixing_time_lower_bound_nonconvex}
	Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy Assumptions \ref{asmp: cluster_diameter} and \ref{asmp: bounded_density} for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D > 0$. Then, for any $r < \sigma/4d$, the following statement holds with probability at least $1 - \delta$: Fix $0 < \epsilon > 1$. Then, for $n$ satisfying:
	\begin{align*}
	\sqrt{3^{d+1}\frac{(\log n + d\log \mu + \log(4/\delta))}{n \nu_d r^d \lambda_{\sigma}}} \leq \epsilon
	\end{align*}
	we have
	\begin{align}
	\label{eqn: inverse_mixing_time_lower_bound_nonconvex}
	1 / \Psi_{n,r}(\Csig[\Xbf]) & \leq \left(d \log \mu + c_{\lambda} \right) \cdot \nonumber \\ 
	& \left(c_1 + c_2 \frac{\Lambda_{\sigma}^4D^{2d}}{\lambda_{\sigma}^4r^{2d}}\left(d \log \mu + c_{\lambda}\right)\right)
	\end{align}
	where $\mu = \log(\frac{2D}{r})$; $c_1$, $c_2$, and $c_3$ are constants which depend only on dimension; and $c_{\lambda} = \log(\Lambda_{\sigma}^2/ \lambda_{\sigma}^2)$. 
\end{theorem}

\begin{remark}
	The proof of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} relies on upper bounding the mixing time using the \emph{conductance} of $G_{n,r}[\Csig[X]]$,
	\begin{equation*}
	\widetilde{\Phi} = \min_{S \subseteq \Csig[\Xbf]} \Phi(S; G_{n,r}[\Csig[X]])
	\end{equation*}

	The factor of $\frac{1}{r^{2d}}$ present in the bound of \eqref{eqn: inverse_mixing_time_lower_bound_nonconvex} is suboptimal. This exponential dependence on $d$ stems from a loose bound on the aforementioned conductance in the proof of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex}.In particular, we assert only that any set $S \subseteq \Csig[X]$ must have $\cut(S; \Csig[\Xbf])$ on the order of $n^2 r^{2d}$, while upper bounding $\vol(S; \Csig[\Xbf])$ by roughly $n^2 r^d$, for a bound on the conductance of order $r^d$. The presence of $r^{2d}$ comes from upper bounding the mixing time by about $1/\widetilde{\Phi}^2$, this latter bound being a variant of classic results on rapid mixing \cite{jerrum89}.
\end{remark}

It is possible to sharpen the dependency on $d$, but at the cost of an additional assumption:
\begin{enumerate}[label=(A\arabic*)]
	\setcounter{enumi}{4}
	\item 
	\label{asmp: convex}
	$\Cset$ is convex.
\end{enumerate}
Additionally, the resulting bound holds only asymptotically.

>>>>

\begin{abstract}
\vskip 0.1 in % RJT: there was no space here and it looked very squished 
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several authors have shown that local methods, such as personalized
PageRank (PPR), have worst-case guarantees for certain graph-based measures of
cluster quality.  In contrast to existing works, we analyze PPR in a traditional
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.
\end{abstract}

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be been computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

Local spectral clustering techinques have been practically very successful
\citep{leskovec2010,andersen2012,gleich2012,mahoney2012,wu2012}, which has led
many authors to develop supporting theory
\citep{spielman2013,andersen2009,gharan2012,zhu2013} that gives worst-case
guarantees on traditional graph-theoretic notions of cluster quality (like
conductance).  In this paper, we adopt a more traditional statistical viewpoint,
and examine what the output of a local clustering algorithm on $\Xbf$ reveals
about the unknown density $f$.  In particular, we examine the ability of the PPR
algorithm to recover \emph{density clusters} of $f$, which are defined as the
connected components of the upper level set $\{x \in \Rd : f(x) \geq \lambda\}$
for some threshold $\lambda > 0$ (a central object of central interest in the
classical statistical literature on clustering, dating back to
\citealt{hartigan1981}).

@@

If $S$ has normalized cut no greater than $\Phi$, and inverse mixing time no
less than $\Psi$, we call it as a \emph{$(\Phi,\Psi)$-cluster}. Both
local \citep{zhu2013} and global \citep{kannan04} spectral algorithms have been
shown to output clusters (or partitions) which approximate the optimal $(\Phi,
\Psi)$-cluster (or partition) for a given graph $G$.\footnote{In the case of
  \citet{kannan04}, the internal connectivity parameter $\phi$ is actually the
  conductance, i.e., the minimum normalized cut within the subgraph $G[S]$. See
  Theorem 3.1 in their paper for details; however, note that $\phi^2 /
  \log(\vol(S)) \leq O(\Psi)$, and so the lower bound on $\phi$ translates to a
  lower bound on $\Psi$.}

@@

For a level $\beta > 0$ and a target volume $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of PPR
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target volume $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return \smash{$\widehat{C} = S_{\beta^*}$}, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

@@

%% RJT: This was already commented out before

Let $\Cbb_f(\lambda)$ denote the connected components of the density upper level
set $\{x \in \Rd: f(x) > \lambda\}$.  For a given density cluster $\Cset \in
\Cbb_f(\lambda)$, we call $\Cset[\Xbf] = \Cset \cap \Xbf$ the \emph{empirical
  density cluster}. Below we define a notion of consistency in density cluster
estimation.

@@

\begin{enumerate}
\item In Section \ref{sec: PhiPsi}, we derive in Theorem \ref{thm:
    conductance_upper_bound} an upper bound on the normalized cut of a
  (thickened) empirical density cluster $\Csig[\Xbf]$, under natural geometric 
  conditions (precluding clusters that are too thin and long).

\item Under largely the same set of geometric conditions, we derive in Theorem
  \ref{thm: inverse_mixing_time_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Csig[\Xbf]$.  We also provide in 
  Theorem \ref{thm: inverse_mixing_time_lower_bound} a tighter lower bound, but
  under more restrictive assumptions (convexity of $\Csig$).

@@

\section{Cluster Quality Critera Bounds for Density Clusters}  
\label{sec: PhiPsi}

\subsection{Assumptions}

In order to provide meaningful bounds on the normalized cut and inverse mixing
time of an empirical density cluster, we must introduce conditions on the
density $f$. Let $B(x,r) = \{y \in \Rd: \norm{y - x} \leq r\}$ be the ball of
radius $r$ centered at $x$.  Given a set $\Aset \subseteq \Rd$ and $\sigma > 0$,
define $\Asig = \Aset + B(0,\sigma) = \{y \in \Rd: \inf_{x \in \Aset} \norm{y - 
    x} \leq \sigma\}$, which we call the $\sigma$-expansion of $\Aset$.

We are now ready to give the conditions, stated with respect to a density
cluster $\Cset \in \Cbb_f(\lambda)$, for some $\lambda > 0$, and an expansion 
parameter $\sigma > 0$.

\begin{enumerate}[label=(A\arabic*)]
\item
\label{asmp: bounded_density}
\emph{Bounded density within cluster:} There are $0 < \lambda_{\sigma} <
\Lambda_{\sigma} < \infty$ such that
$$
\lambda_{\sigma} = \inf_{x \in \Csig} f(x) \leq \sup_{x \in \Csig} f(x) \leq
\Lambda_{\sigma}.
$$
% and 
% \begin{equation*}
% \frac{\diam \Asig}{\sigma} \leq \mu
% \end{equation*}
% where $\diam \Asig = \sup \set{d(x,y) : x,y \in \Asig}$

\item 
\label{asmp: low_noise_density}
\emph{Low noise density:} There exist $\gamma, c_1 > 0$ such that for all $x
\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
$$
  \inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \dist(x, \Csig)^{\gamma},
$$
where \smash{$\dist(x,\Aset) = \inf_{x_0 \in \Aset} \norm{x - x_0}$}.
% RJT: should we change this to c_0?  c_1 seems to be used below and throughout
% for other constants, and c_0 is not used

\item
\label{asmp: cluster_separation}
\emph{Cluster separation:}
For all $\Cset' \in \Cbb_f(\lambda)$,
$$
\dist(\Csig,\Csig') > \sigma,
$$
where \smash{$\dist(\Aset,\Aset') = \inf_{x \in \Aset} \dist(x,\Aset')$}.

\item
\label{asmp: cluster_diameter}
\emph{Cluster diameter:}
There exists $D < \infty$ such that for all $x, x' \in \Csig$,
$$
\norm{x - x'} \leq D.
$$
\end{enumerate}

Note that $\sigma$ plays several roles here, precluding arbitrarily narrow
clusters and long clusters in \ref{asmp: bounded_density} and \ref{asmp:
  cluster_diameter}, flat densities around the level set in \ref{asmp: 
  low_noise_density}, and poorly separated clusters in \ref{asmp:
  cluster_separation}.

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} and \ref{asmp:  
cluster_separation} combined are similar to the $(\sigma,\epsilon)$-saliency of
\citet{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citet{singh2009} (as well as
many other works on density clustering and level set estimation.) It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as PPR.

In the next several subsections, we will derive bounds on the cluster quality
criteria evaluated on (thickened) density clusters. For notational simplicity,
hereafter for $S \subseteq V$, we will denote $\Phi(S; G_{n,r})$ by
$\Phi_{n,r}(S)$, and likewise with $\Psi(S; G_{n,r})$ and $\Psi_{n,r}(S)$. We
will also use $\nu$ for Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ for 
the measure of the unit ball $B=B(0,1)$. 
% RJT: you had B(0,d) before ... just check this should have been B(0,1)

\subsection{Upper Bound on Normalized Cut}

We begin by upper bounding the normalized cut of $\Cset_\sigma[\Xbf]$.  (In the
theorem, the upper bound on the density in Assumption \ref{asmp: 
  bounded_density} will not actually be needed, so we omit the parameter
$\Lambda_\sigma>0$ from the theorem statement.)

\begin{theorem}
\label{thm: conductance_upper_bound}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation}, for some 
$\sigma, \lambda_{\sigma}, c_1, \gamma > 0$. Then for any $0 < r <
\sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and
\begin{equation}
\label{eqn: conductance_sample_complexity}
n \geq \frac{9\log(2/\delta)}{\epsilon^2}\left(\frac{1}
  { \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
\end{equation}
we have
\begin{equation}
  \label{eqn: conductance_additive_error_bound}
  \frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq 4 c_{\sigma} d
  \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
    \frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
\end{equation}
with probability at least $1-\delta$, where $c_{\sigma} = 1 / \sigma$. 
\end{theorem}

\begin{remark}
The proof of Theorem \ref{thm: conductance_upper_bound}, along with all other
proofs in this paper, can be found in the supplementary document. The key 
idea is that for any $x \in \Cset$, the simple (possibly loose) fact
$B(x,\sigma) \subseteq \Csig$ translates into the upper bound $\nu(\Cset_{\sigma
  + r}) \leq (1 + 2d r /\sigma)\nu(\Csig)$. We leverage \ref{asmp:
  low_noise_density} to find a corresponding bound on the weighted volume,
then apply standard concentration inequalities to convert from population-
to sample-based results.  
\end{remark}

\begin{remark}
The inequality in \eqref{eqn: conductance_additive_error_bound} is almost
tight. Specifically, choosing $\Asig = B(0,\sigma)$ and
$$
f(x) = \lambda - \dist(x,\Asig)^{\gamma},
$$
% RJT: is this correct (is this what you meant)?  Original formula is below
% $$
% f(x)  = \begin{cases}
%   \lambda &\text{for $x \in \Asig$}, \\
%   \lambda - \dist(x,\Asig)^{\gamma} & \text{ for $0 < \dist(x,\Asig) < r$}, 
% \end{cases}
% $$
we have that for $n$ on the order of the lower bound in \eqref{eqn: 
  conductance_sample_complexity}, 
\begin{equation*}
  \frac{\Phi_{n,r}(\Asig[\mathbf{X}])}{r} \geq c \frac{(\lambda -
    \frac{r^{\epsilon+1}}{\epsilon+1})}{\lambda} - \epsilon,
\end{equation*}
with probability at least $1 - \delta$, for some constant $c$. (Note that the
factor of $1 / \sigma$ in $c_{\sigma}$ is not replicated above.)
\end{remark}

\subsection{Lower Bound on Inverse Mixing Time}

We lower bound the inverse mixing time of $\Csig[\Xbf]$.

\begin{theorem}
\label{thm: inverse_mixing_time_lower_bound_nonconvex}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D > 0$. Then for any $0 <
r < \sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and $n$ satisfying 
$$
\sqrt{3^{d+1}\frac{(\log n + d\log \mu + \log(4/\delta))}{n \nu_d r^d
    \lambda_{\sigma}}} \leq \epsilon, 
$$
where \smash{$\mu = \log(\frac{2D}{r})$}, we have
\begin{multline}
\label{eqn: inverse_mixing_time_lower_bound_nonconvex}
\Psi_{n,r}(\Csig[\Xbf]) \geq \left(d \log \mu + c_{\lambda} \right)^{-1} \cdot \\
\left(c_1 + c_2 \frac{\Lambda_{\sigma}^4D^{2d}}{\lambda_{\sigma}^4r^{2d}}\left(d
    \log \mu + c_{\lambda}\right)\right)^{-1},
\end{multline}
% RJT: confused ... is this correct?
with probability at least $1 - \delta$, where $c_1,c_2,c_3>0$ are constants
depending only on the dimension $d$, and $c_{\lambda} = \log(\Lambda_{\sigma}^2/
\lambda_{\sigma}^2)$.   
\end{theorem}
% RJT: saying that c_1,c_2,c_3 depend on the dimension is a little odd/weak.
% c_2 multiplyies, e.g., a factor of d \log \mu.  So if it's d^{10}, then this
% will swamp the linear factor of ....

\begin{remark}
The proof of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} relies
on upper bounding the mixing time using the \emph{conductance} of $G_{n,r}[\Csig[X]]$, 
$$
\widetilde{\Phi} = \min_{S \subseteq \Csig[\Xbf]} \Phi(S; G_{n,r}[\Csig[X]]).
$$
The factor $1/r^{2d}$ in the lower bound in \eqref{eqn:
  inverse_mixing_time_lower_bound_nonconvex} is suboptimal. This exponential
dependence on $d$ stems from a loose bound on the aforementioned
conductance. In particular, we assert only that any set $S \subseteq \Csig[X]$
must have $\cut(S; \Csig[\Xbf])$ on the order of $n^2 r^{2d}$, while upper bounding
$\vol(S; \Csig[\Xbf])$ by roughly $n^2 r^d$, for a bound on the conductance of
order $r^d$. The presence of $r^{2d}$ comes from upper bounding the mixing time
by about \smash{$1/\widetilde{\Phi}^2$}, this being a variant of classic results
on rapid mixing \citep{jerrum89}.  
\end{remark}

\subsection{Tigher Lower Bound Under Convexity}

It is possible to sharpen the dependency on $d$ in 
\label{eqn: inverse_mixing_time_lower_bound_nonconvex}, but at the cost of an 
additional assumption: 
\begin{enumerate}[label=(A\arabic*)]
  \setcounter{enumi}{4}
\item 
  \label{asmp: convex}
  $\Cset$ is convex.
\end{enumerate}
% RJT: do we need C to satisfy this, or Csig?  Not sure I know why the former
% matters ...

With \ref{asmp: convex} in place, we now give a tighter lower bound on the
inverse mixing time of $C_\sigma[\Xbf]$ (albeit one that holds only
asymptotically).

\begin{theorem}
\label{thm: inverse_mixing_time_lower_bound}
Assume the conditions of Theorem \ref{thm:
  inverse_mixing_time_lower_bound_nonconvex}, and additionally, let $\Cset$
satisfy \ref{asmp: convex}. Then, 
for any $r < \sigma/4d$, the following statement holds with probability 1: 
\begin{align}
\label{eqn: inverse_mixing_time_lower_bound}
  \liminf_{n \to \infty} \, &\Psi_{n,r}(\Csig[\Xbf]) \geq \nonumber \\ 
&\quad c_1 \frac{\Lambda_{\sigma}^8}{\lambda_{\sigma}^8} (\log \mu +
  c_{\lambda}) \biggl( c_{\lambda}(d^3 \log\mu + c_2) +{} \nonumber \\  
&\quad \frac{d^2D^2}{r^2} \bigl( \log d + c_2
  \frac{\Lambda_{\sigma}^{2}}{\lambda_{\sigma}^{2}}c_{\lambda} +{} \nonumber \\  
&\quad c_3  + \log \log \mu \bigr) \biggr) + c_d \frac{o_r(1)}{r^2},
\end{align}
% RJT: confused ... is this correct?
where \smash{$\mu = \log(\frac{2D}{r})$}, $c_1,c_2,c_3>0$ are universal
constants, $c_d$ is constant in $r$ but depends on dimension $d$, and
$c_{\lambda} = \log(\Lambda_{\sigma}^2/ \lambda_{\sigma}^2)$.  
% Additionally, $\lim_{r \to 0} o_r(1) = 0$.  
\end{theorem}
% RJT: this was a lower bound on 1/\Psi and I changed it to \Psi ... just
% checking
